---
title: "chapter6.Rmd"
author: "Elmeri Santala"
date: "12/9/2019"
output: html_document
---

# Exercise 6 / Analysis of longitudinal data
## Datawrangling
Code for this weeks datawrangling is here:
https://github.com/elmeris/IODS-project/blob/master/data/meet_and_repeat.R

So lets start!

```{r}
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
```

### BPRS
The Brief Psychiatric Rating Scale (BPRS) is a rating scale which a clinician or researcher may use to measure psychiatric symptoms such as depression, anxiety, hallucinations and unusual behaviour. Each symptom is rated 1-7 and depending on the version between a total of 18-24 symptoms are scored (i.e. Somatic concern, Anxiety, Depression, Suicidality). The scale is one of the oldest, most widely used scales to measure psychotic symptoms and was first published in 1962.

This data includes 40 rows and 11 variables. 
```Treatment```: treatment 1 or 2. 
```Subject```: patient study id
```Week```: study week 0-8. 

```{r}
colnames(BPRS)
str(BPRS)
summary(BPRS)
```

```{r}
library(dplyr)
library(tidyr)
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))
```

```{r}
glimpse(BPRSL)
str(BPRSL)
dim(BPRSL)
names(BPRSL)
```

### RATS
The RATS data is from a study conducted in three groups of rats. These three groups were put on different diets, and each animal's body weight (grams) was recorded repeatedle(approximately weekly, expect in week seven when two recordings were taken) over 9-week period. The question of most interest is whether the growth profiles of the three groups differ.

Data set has 16 rows and 13 variables. 
```{r}
colnames(RATS)
str(RATS)
summary(RATS)
```

```{r}
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
RATSL <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) %>%
  mutate(Time = as.integer(substr(WD,3,4))) 
```

```{r}
glimpse(RATSL)
str(RATSL)
dim(RATSL)
names(RATSL)
```

Data sets are now wrangled lets start!

## Analyses of Chapter 8 of MABS using the RATS data

```{r}
library(MASS)
library(tidyr)
library(stringr)
library(ggplot2)
library(GGally)
library(corrplot)
library(tidyverse)
```

```{r}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```
**Describtion**
Graph shows us that rats in the group 2 and 3 had higher weight from the start and gained more weight during study that rats in the group 1.
Rats in the group 2 and 3 have more variety in the strating weight and end weight.
Inside the all groups the are rats which gained more weight during 8-weeks than others.

Next lets standardise the RATS data

```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
```

```{r}
glimpse(RATSL)
```

```{r}
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized Weight")
```

```{r}
n <- RATSL$Time %>% unique() %>% length()
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()
```

```{r}
glimpse(RATSS)
```

```{r}
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```
**Describtion**
Standardised RATS data (RATSS) has 33 rows and 4 variables. 
Variables are ```Group``` ```Time``` ```mean``` ```se```.

The graph above show us that the rats in the all three groups have gained weight during the study. Group 1 has lowest average strating weight and end weight. Group 3 has the highest average strating weight and and weight. 

Lets create the summary data RATSL8S!

```{r}
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
```

```{r}
glimpse(RATSL8S)
```

```{r}
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-8")
```
**Describtion**
This graph shows us same result that we have seen already before. Also there is some individul rat in each group which has strongly different weight than others.

Lets filter "outlier" from mean under 250 and create a new graph!
```{r}
RATSL8S1 <- RATSL8S %>%
  filter(mean > 250)
```

```{r}
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-8")
```

Lets filter also "outlier" from mean over 550 and create a new graph!

```{r}
RATSL8S2 <- RATSL8S1 %>%
  filter(mean < 550)
```

```{r}
ggplot(RATSL8S2, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-8")
```
**Describtion**
As we can see now the boxplots are different in group 1 and 2 after filtering outliers from both ends of the scale.

Lets perform anova analysis to this data!

```{r}
fit <- lm(RATSL8S2$mean ~ RATSL8S2$Group, data = RATSL8S2)
anova(fit)
```
**Describtion**
Table shows us that P-value is low so there is statistically significant difference in the mean weight between the groups.

## Analyses of Chapter 9 of MABS using the BPRS data.

Lets check the dimensions of the data first.

```{r}
dim(BPRSL)
```
There is 360 rows and 5 variables in the data.

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, group = interaction(subject, treatment))) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "week", breaks = seq(0, 8, 4)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
```
**Description**
Graph above show us each patients BPRS scores during the study. Lines are different in the group 1 and 2.

Lets create a regression model BPRS_reg!

```{r}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
```

```{r}
summary(BPRS_reg)
```
**Describtion**
This model shows us that week is statistically significant effect to BPRS (<2e-16 ***) and treatment 2 has not (0.661).

Lets create a random intercept model!


```{r}
library(lme4)
```

```{r}
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
```

```{r}
summary(BPRS_ref)
```

Lets create a random intercept and random slope model!

```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
```

```{r}
summary(BPRS_ref1)
```

Lets perform an ANOVA test on the two models!
```{r}
anova(BPRS_ref1, BPRS_ref)
```
**Describtion**
Table shows us that chi square is 7.27, and the associated p-value is 0.026.
The lower the Chi square value is the better the fit of the model against the comparison model.

Lets create a random intercept and random slope model!

```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
```

```{r}
summary(BPRS_ref2)
```

```{r}
anova(BPRS_ref2, BPRS_ref1)
```
**Describtion**
Table shows us that chi square is now 3.17, and the associated p-value is 0.076.

Lets create a vector of the fitted values, a new column fitted to BPRSL and draw the plot of BPRSL!

```{r}
Fitted <- fitted(BPRS_ref2)
```

```{r}
BPRSL <- BPRSL %>%
  mutate(Fitted)
```

```{r}
ggplot(BPRSL, aes(x = week, y = Fitted, group = interaction(subject, treatment))) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "week", breaks = seq(0, 8, 4)) +
  scale_y_continuous(name = "Fitted bprs") +
  theme(legend.position = "top")
```
**Describtion**
Graph above show us that the fitted values of BPRS changes lineary. But, as we know from same kind graph earlier that in real life in the study the BPRS values didn't change lineary.
