# Exercise2: Analysis
Code for data wrangling: https://github.com/elmeris/IODS-project/blob/master/data/create_learning2014.R

**reading data**
```{r}
learning2014 <- read.csv("~/IODS-project/data/learning2014.csv")
str(learning2014)
head(learning2014)
```
```{r}
dim(learning2014)
```
There is seven variables and 166 observations.

**Graphical overview**

```{r fig2, fig.dim=c(10,10), results='hide', message=FALSE}
# access the GGally and ggplot2 libraries
library(GGally)
library(ggplot2)

# create a more advanced plot matrix with ggpairs()
p <- ggpairs(learning2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

# draw the plot
p
```
Highest correlation is between points and attitude. All distributions looks even but age is converted to the right.

**Regression model**

The explanatory variables are "surface", "attitude" and "strategy" 

*Scatter plot*
```{r}
qplot(attitude, points, data = learning2014) + geom_smooth(method = "lm")
```
*a linear model*
```{r}
model <- lm(points ~1, data = learning2014)
```
*Summary of the model*
```{r}
summary(model)
```

*Regression model with multiple explanatory variables*
```{r}
my_model2 <- lm(points ~ attitude + stra + surf, data = learning2014)
```

*Summary of the model2*
```{r}
summary(my_model2)
```

Lets move surf out because the p-vaöue is the highest.

*New model*
```{r}
my_model3 <- lm(points ~ attitude + stra, data = learning2014)
```
*summary*
```{r}
summary(my_model3)
```
Attitude is significant variable. Strategy is not statistically significant.
R-squared:the higher the R-squared, the better the model fits your data. R-squared is always between 0 and 100%.
Multiple R-squared:  0.2048. Means the model fits the data quite well.

**Diagnostic plot**
```{r}
par(mfrow = c(2,2))
plot(my_model3, which = c(1,2,5))
```
QQ plot: model is normally distributed